---
title: "LGBTQI+ Hate Crime Analysis — 01: Data Pipeline"
author: "Guilherme Arpi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_knit$set(
  root.dir = if (basename(getwd()) == "R") ".." else "."
)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

## Overview

This notebook ingests the raw FBI Uniform Crime Reporting (UCR) hate crime dataset, filters for LGBTQI+-motivated incidents, engineers features, and produces three analysis-ready CSV files used by all downstream scripts.

**Inputs**

| File | Description |
|------|-------------|
| `data/hate_crime.csv` | FBI UCR hate crime statistics, 1991–2024 (~265k rows) |
| `data/gini_index_1991_2024.csv` | Year-specific Gini coefficient by state (built by `build_gini_dataset.py`) |
| `data/total_population_1991_2024.csv` | Total state population, Census intercensal estimates (built by `build_population_datasets.py`) |
| `data/lgbt_population_2012_2024.csv` | LGBT adult population % by state, 2012–2024 (built by `build_population_datasets.py`) |
| `data/state_law_panel.csv` | Time-varying hate crime law coverage per state × year (built by `build_law_panel.py`) |

**Outputs**

| File | Description |
|------|-------------|
| `data/lgbtq_hate_crimes_full.csv` | All LGBTQI+ incidents, 1991–2024 (~45k rows) |
| `data/lgbtq_hate_crimes_enriched.csv` | Incident-level data (2010–2024) joined with Gini, population, and law panel — ML-ready |
| `data/state_year_aggregate.csv` | State × year summary with two-track normalised rates — Tableau-ready |

---

## Setup

```{r libraries}
library(tidyverse)
library(lubridate)
```

```{r paths}
BASE_DIR <- if (basename(getwd()) == "R") dirname(getwd()) else getwd()  # works from repo root or from R/
RAW_FBI      <- file.path(BASE_DIR, "data", "hate_crime.csv")
RAW_GINI     <- file.path(BASE_DIR, "data", "gini_index_1991_2024.csv")
RAW_POP_TOT  <- file.path(BASE_DIR, "data", "total_population_1991_2024.csv")
RAW_POP_LGBT <- file.path(BASE_DIR, "data", "lgbt_population_2012_2024.csv")
RAW_LAW_PANEL <- file.path(BASE_DIR, "data", "state_law_panel.csv")
DATA_DIR     <- file.path(BASE_DIR, "data")

INCLUDE_2026_SRS <- FALSE
SRS_FILE <- file.path(DATA_DIR, "2026_HC_NATIONAL_MASTER_FILE.txt")

dir.create(DATA_DIR, showWarnings = FALSE, recursive = TRUE)
```

---

## 1. Load & Filter FBI Data

The raw FBI UCR file covers all reported hate crime incidents across all bias categories (race, religion, national origin, sexual orientation, gender identity, disability, etc.). We filter to LGBTQI+-motivated incidents using keyword matching on the `BIAS_DESC` field.

The 2024 file ships with lowercase column names; `rename_with(str_to_upper)` normalises to uppercase for consistency across vintages.

```{r load-fbi}
fbi <- read_csv(RAW_FBI, show_col_types = FALSE) |>
  rename_with(str_to_upper) |>
  rename(any_of(c(
    POPULATION_GROUP_DESC = "POPULATION_GROUP_DESCRIPTION"
  )))

cat(sprintf("Raw rows: %s\n", format(nrow(fbi), big.mark = ",")))

# Filter to LGBTQI+ bias motivations
lgbt_terms <- c("gay", "lesbian", "bisexual", "transgender",
                "gender non-conform", "sexual orient", "gender identity")
pattern    <- paste(lgbt_terms, collapse = "|")

df <- fbi |>
  filter(str_detect(str_to_lower(BIAS_DESC), pattern)) |>
  mutate(across(everything(), as.character))

cat(sprintf("LGBTQI+ rows: %s (%.1f%% of total)\n",
            format(nrow(df), big.mark = ","),
            100 * nrow(df) / nrow(fbi)))
```

---

## 2. Clean & Type-Cast

Date parsing uses `coalesce(ymd(), dmy())` to handle both the ISO format (`YYYY-MM-DD`) used in newer FBI files and the legacy `DD-MMM-YYYY` format used in older ones.

```{r clean}
df <- df |>
  mutate(
    INCIDENT_DATE          = coalesce(ymd(INCIDENT_DATE), dmy(INCIDENT_DATE)),
    YEAR                   = as.integer(DATA_YEAR),
    MONTH                  = month(INCIDENT_DATE),
    MONTH_NAME             = month(INCIDENT_DATE, label = TRUE, abbr = TRUE) |> as.character(),
    VICTIM_COUNT           = as.integer(VICTIM_COUNT),
    TOTAL_OFFENDER_COUNT   = as.integer(TOTAL_OFFENDER_COUNT),
    ADULT_VICTIM_COUNT     = as.integer(replace_na(ADULT_VICTIM_COUNT,    "0")),
    JUVENILE_VICTIM_COUNT  = as.integer(replace_na(JUVENILE_VICTIM_COUNT, "0")),
    JUVENILE_OFFENDER_COUNT = as.integer(replace_na(JUVENILE_OFFENDER_COUNT, "0")),
    MULTIPLE_OFFENSE_FLAG  = as.integer(MULTIPLE_OFFENSE == "M"),
    MULTIPLE_BIAS_FLAG     = as.integer(MULTIPLE_BIAS    == "M"),
  )
```

---

## 3. Feature Engineering

Four helper functions derive simplified categorical features from raw text fields. These make downstream analysis and visualisation cleaner.

```{r features}
# Simplified bias category
classify_bias <- function(bias) {
  b <- str_to_lower(bias)
  case_when(
    str_detect(b, "gender non-conform")               ~ "Anti-Gender Non-Conforming",
    str_detect(b, "transgender")                       ~ "Anti-Transgender",
    str_detect(b, "bisexual")                          ~ "Anti-Bisexual",
    str_detect(b, "lesbian") & !str_detect(b, "gay")  ~ "Anti-Lesbian",
    str_detect(b, "lesbian") & str_detect(b, "gay")   ~ "Anti-LGBTQ (General)",
    str_detect(b, "gay")                               ~ "Anti-Gay (Male)",
    TRUE                                               ~ "Anti-LGBTQ (Other)"
  )
}

# Offense severity (0 = non-violent, 1 = simple assault, 2 = serious violent)
violent_terms <- c("Murder", "Rape", "Aggravated Assault", "Simple Assault",
                   "Robbery", "Arson", "Human Trafficking")

severity_tier <- function(offense) {
  case_when(
    str_detect(offense, "Murder|Rape")                    ~ 2L,
    str_detect(offense, "Aggravated Assault|Robbery|Arson") ~ 2L,
    str_detect(offense, "Simple Assault")                 ~ 1L,
    TRUE                                                  ~ 0L
  )
}

# Season
season_from_month <- function(m) {
  case_when(
    m %in% c(12, 1, 2)   ~ "Winter",
    m %in% c(3, 4, 5)    ~ "Spring",
    m %in% c(6, 7, 8)    ~ "Summer",
    m %in% c(9, 10, 11)  ~ "Fall",
    TRUE                  ~ NA_character_
  )
}

# Agency and location simplifications
simplify_agency <- function(a) {
  case_when(
    str_detect(a, "City")       ~ "City Police",
    str_detect(a, "County")     ~ "County Sheriff",
    str_detect(a, "University") ~ "University/College PD",
    str_detect(a, "State")      ~ "State Police",
    TRUE                        ~ "Other"
  )
}

simplify_location <- function(loc) {
  case_when(
    str_detect(loc, "Residence|Home")                     ~ "Residence/Home",
    str_detect(loc, "Street|Highway|Road|Alley|Sidewalk") ~ "Street/Highway",
    str_detect(loc, "School|College")                     ~ "School/Campus",
    str_detect(loc, "Church|Synagogue|Mosque|Temple")     ~ "Religious Institution",
    str_detect(loc, "Bar|Restaurant")                     ~ "Bar/Restaurant",
    str_detect(loc, "Park")                               ~ "Park/Outdoor",
    TRUE                                                  ~ "Other"
  )
}

df <- df |>
  mutate(
    BIAS_CATEGORY      = classify_bias(BIAS_DESC),
    IS_VIOLENT         = as.integer(str_detect(OFFENSE_NAME,
                           paste(violent_terms, collapse = "|"))),
    SEVERITY_TIER      = severity_tier(OFFENSE_NAME),
    SEASON             = season_from_month(MONTH),
    JUVENILE_INVOLVED  = as.integer(JUVENILE_VICTIM_COUNT > 0 |
                                    JUVENILE_OFFENDER_COUNT > 0),
    AGENCY_TYPE_SIMPLE = simplify_agency(AGENCY_TYPE_NAME),
    LOCATION_SIMPLE    = simplify_location(LOCATION_NAME)
  )
```

---

## 4. Export Full Dataset (1991–2024)

The full dataset retains all LGBTQI+ incidents across the entire 34-year window. It is the primary input for trend analysis and the EDA script.

```{r export-full}
full_cols <- c(
  "INCIDENT_ID", "YEAR", "MONTH", "MONTH_NAME", "SEASON",
  "STATE_NAME", "STATE_ABBR", "REGION_NAME", "DIVISION_NAME",
  "BIAS_DESC", "BIAS_CATEGORY",
  "OFFENSE_NAME", "IS_VIOLENT", "SEVERITY_TIER",
  "VICTIM_COUNT", "TOTAL_OFFENDER_COUNT",
  "ADULT_VICTIM_COUNT", "JUVENILE_VICTIM_COUNT",
  "JUVENILE_INVOLVED",
  "LOCATION_NAME", "LOCATION_SIMPLE",
  "OFFENDER_RACE", "OFFENDER_ETHNICITY",
  "AGENCY_TYPE_NAME", "AGENCY_TYPE_SIMPLE",
  "POPULATION_GROUP_DESC",
  "MULTIPLE_OFFENSE_FLAG", "MULTIPLE_BIAS_FLAG",
  "VICTIM_TYPES"
)

full_out <- file.path(DATA_DIR, "lgbtq_hate_crimes_full.csv")
df |> select(all_of(full_cols)) |> write_csv(full_out)

cat(sprintf("Saved → lgbtq_hate_crimes_full.csv\n  %s rows | Years: %d–%d\n",
            format(nrow(df), big.mark = ","),
            min(df$YEAR), max(df$YEAR)))
```

---

## 5. Load Reference Datasets

Three reference datasets provide the denominators and covariates for normalisation and regression:

- **Gini index** (state × year, 1991–2024): Frank-Sommeiller-Price historical series (1991–2005) bridged to ACS estimates (2006–2024), with linear interpolation for intercensal years.
- **Total population** (state × year, 1991–2024): Census decennial anchors (2000, 2010, 2020) with linear interpolation.
- **LGBT population** (state × year, 2012–2024 only): Williams Institute / Gallup / BRFSS state-level estimates. **Pre-2012 data does not exist** — backward interpolation was deliberately avoided.
- **State law panel** (state × year, 1991–2024): Time-varying hate crime law coverage from the Movement Advancement Project. `LAW_LEVEL` = 0 (no coverage), 1 (sexual orientation only), 2 (SO + gender identity).

```{r load-reference}
gini_panel <- read_csv(RAW_GINI, show_col_types = FALSE) |>
  rename(STATE_NAME = State, DATA_YEAR = Year, GINI_COEFFICIENT = GINI_Index)

pop_total <- read_csv(RAW_POP_TOT, show_col_types = FALSE) |>
  rename(STATE_NAME = State, DATA_YEAR = Year, TOTAL_POPULATION = Total_Population)

pop_lgbt <- read_csv(RAW_POP_LGBT, show_col_types = FALSE) |>
  rename(STATE_NAME = State, DATA_YEAR = Year, LGBT_PCT = LGBT_Pct)

law_panel <- read_csv(RAW_LAW_PANEL, show_col_types = FALSE) |>
  rename(STATE_NAME = State, DATA_YEAR = Year)

cat(sprintf(
  "Gini panel:       %s rows (%d–%d)\nTotal population: %s rows (%d–%d)\nLGBT population:  %s rows (%d–%d) — 2012+ only\nLaw panel:        %s rows (%d–%d)\n",
  format(nrow(gini_panel), big.mark=","), min(gini_panel$DATA_YEAR), max(gini_panel$DATA_YEAR),
  format(nrow(pop_total),  big.mark=","), min(pop_total$DATA_YEAR),  max(pop_total$DATA_YEAR),
  format(nrow(pop_lgbt),   big.mark=","), min(pop_lgbt$DATA_YEAR),   max(pop_lgbt$DATA_YEAR),
  format(nrow(law_panel),  big.mark=","), min(law_panel$DATA_YEAR),  max(law_panel$DATA_YEAR)
))
```

---

## 6. State × Year Aggregate

The aggregate table is the primary Tableau data source. One row per state per year, with:

- **Track 1** — `RATE_PER_100K_TOTAL`: incidents per 100k total population. Available for all years 1991–2024.
- **Track 2** — `RATE_PER_100K_LGBT`: incidents per 100k LGBT adult residents. Available 2012–2024 only; `NA` for earlier years by design.
- **`LAW_LEVEL`**: the law coverage in effect for that specific state × year (0/1/2).

```{r state-year-aggregate}
state_year_counts <- df |>
  mutate(DATA_YEAR = as.integer(YEAR)) |>
  group_by(DATA_YEAR, STATE_NAME) |>
  summarise(
    REGION_NAME   = first(REGION_NAME),
    INCIDENTS     = n(),
    TOTAL_VICTIMS = sum(as.integer(VICTIM_COUNT), na.rm = TRUE),
    VIOLENT_COUNT = sum(as.integer(IS_VIOLENT),   na.rm = TRUE),
    AVG_SEVERITY  = mean(as.numeric(SEVERITY_TIER), na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(VIOLENT_PCT = round(VIOLENT_COUNT / INCIDENTS * 100, 1))

state_year <- state_year_counts |>
  left_join(gini_panel, by = c("DATA_YEAR", "STATE_NAME")) |>
  left_join(pop_total,  by = c("DATA_YEAR", "STATE_NAME")) |>
  left_join(pop_lgbt,   by = c("DATA_YEAR", "STATE_NAME")) |>
  left_join(law_panel |> select(STATE_NAME, DATA_YEAR, LAW_LEVEL, HC_LAW_SO, HC_LAW_GI),
            by = c("DATA_YEAR", "STATE_NAME")) |>
  mutate(
    RATE_PER_100K_TOTAL = if_else(
      !is.na(TOTAL_POPULATION) & TOTAL_POPULATION > 0,
      round((INCIDENTS / TOTAL_POPULATION) * 100000, 4), NA_real_
    ),
    LGBT_POPULATION_EST = if_else(
      !is.na(LGBT_PCT) & !is.na(TOTAL_POPULATION),
      round(TOTAL_POPULATION * (LGBT_PCT / 100)), NA_real_
    ),
    RATE_PER_100K_LGBT = if_else(
      !is.na(LGBT_POPULATION_EST) & LGBT_POPULATION_EST > 0,
      round((INCIDENTS / LGBT_POPULATION_EST) * 100000, 4), NA_real_
    )
  )

write_csv(state_year, file.path(DATA_DIR, "state_year_aggregate.csv"))
cat(sprintf("Saved → state_year_aggregate.csv\n  %s rows | Years: %d–%d\n",
            format(nrow(state_year), big.mark=","),
            min(state_year$DATA_YEAR), max(state_year$DATA_YEAR)))
```

---

## 7. Enriched Dataset (2010–2024)

The enriched dataset joins each individual incident with its year-specific Gini coefficient, population denominators, and law coverage level. This is the primary input for the machine learning models.

Unlike the full dataset (1991–2024), the enriched dataset is restricted to 2010+ because the Gini and population reference data become sparse before that window, and the ML models benefit from more complete covariate coverage.

```{r enriched-dataset}
enriched <- df |>
  filter(YEAR >= 2010) |>
  mutate(DATA_YEAR = as.integer(YEAR)) |>
  left_join(gini_panel, by = c("DATA_YEAR", "STATE_NAME")) |>
  left_join(pop_total,  by = c("DATA_YEAR", "STATE_NAME")) |>
  left_join(pop_lgbt,   by = c("DATA_YEAR", "STATE_NAME")) |>
  left_join(law_panel |> select(STATE_NAME, DATA_YEAR, LAW_LEVEL, HC_LAW_SO, HC_LAW_GI),
            by = c("STATE_NAME", "DATA_YEAR")) |>
  mutate(
    GINI_COEFFICIENT = as.numeric(GINI_COEFFICIENT),
    HAS_LAW          = as.integer(LAW_LEVEL > 0),
    FULL_PROTECTION  = as.integer(LAW_LEVEL == 2),
    LGBT_POPULATION  = if_else(
      !is.na(LGBT_PCT) & !is.na(TOTAL_POPULATION),
      round(TOTAL_POPULATION * (LGBT_PCT / 100)), NA_real_
    ),
    RATE_PER_100K_TOTAL = if_else(
      !is.na(TOTAL_POPULATION) & TOTAL_POPULATION > 0,
      round((1 / TOTAL_POPULATION) * 100000, 8), NA_real_
    ),
    RATE_PER_100K_LGBT = if_else(
      !is.na(LGBT_POPULATION) & LGBT_POPULATION > 0,
      round((1 / LGBT_POPULATION) * 100000, 8), NA_real_
    )
  )

write_csv(enriched, file.path(DATA_DIR, "lgbtq_hate_crimes_enriched.csv"))
cat(sprintf("Saved → lgbtq_hate_crimes_enriched.csv\n  %s rows | Years: %d–%d\n",
            format(nrow(enriched), big.mark = ","),
            min(enriched$DATA_YEAR, na.rm = TRUE),
            max(enriched$DATA_YEAR, na.rm = TRUE)))
```

---

## Pipeline Summary

```{r summary}
cat("✓ Pipeline complete. Files written to /data/\n\n")
cat(sprintf("  lgbtq_hate_crimes_full.csv      — %s rows, 1991–%d\n",
            format(nrow(df), big.mark=","), max(df$YEAR)))
cat(sprintf("  lgbtq_hate_crimes_enriched.csv  — %s rows, 2010–%d\n",
            format(nrow(enriched), big.mark=","),
            max(enriched$DATA_YEAR, na.rm=TRUE)))
cat(sprintf("  state_year_aggregate.csv        — %s rows, %d–%d (Tableau-ready)\n",
            format(nrow(state_year), big.mark=","),
            min(state_year$DATA_YEAR), max(state_year$DATA_YEAR)))
```

---

## Appendix: 2026 SRS Master File Parser

The FBI distributes a fixed-width SRS National Master File for agencies still on the legacy Summary Reporting System. As of 2024, most agencies report via NIBRS, so this file captures only a fraction of national incidents. Set `INCLUDE_2026_SRS <- TRUE` at the top of this notebook to append partial 2026 records to the full dataset.

```{r srs-lookup-tables}
srs_state_lookup <- tribble(
  ~code, ~abbr, ~name,
  "01","AL","Alabama",      "02","AK","Alaska",      "03","AR","Arkansas",
  "04","AZ","Arizona",      "05","CA","California",  "06","CO","Colorado",
  "07","CT","Connecticut",  "08","DE","Delaware",    "09","DC","District of Columbia",
  "10","FL","Florida",      "11","GA","Georgia",     "12","ID","Idaho",
  "13","IL","Illinois",     "14","IN","Indiana",     "15","IA","Iowa",
  "16","KS","Kansas",       "17","KY","Kentucky",    "18","LA","Louisiana",
  "19","ME","Maine",        "20","MD","Maryland",    "21","MA","Massachusetts",
  "22","MI","Michigan",     "23","MN","Minnesota",   "24","MS","Mississippi",
  "25","MO","Missouri",     "26","MT","Montana",     "27","NE","Nebraska",
  "28","NV","Nevada",       "29","NH","New Hampshire","30","NJ","New Jersey",
  "31","NM","New Mexico",   "32","NY","New York",    "33","NC","North Carolina",
  "34","ND","North Dakota", "35","OH","Ohio",        "36","OK","Oklahoma",
  "37","OR","Oregon",       "38","PA","Pennsylvania","39","RI","Rhode Island",
  "40","SC","South Carolina","41","SD","South Dakota","42","TN","Tennessee",
  "43","TX","Texas",        "44","UT","Utah",        "45","VT","Vermont",
  "46","VA","Virginia",     "47","WA","Washington",  "48","WV","West Virginia",
  "49","WI","Wisconsin",    "50","AK","Alaska",      "51","WY","Wyoming"
)

nibrs_offense_map <- c(
  "09A"="Murder/Non-Negligent Homicide", "13A"="Aggravated Assault",
  "13B"="Simple Assault",               "13C"="Intimidation",
  "120"="Robbery",                       "200"="Arson",
  "290"="Destruction/Damage/Vandalism",  "11A"="Forcible Rape"
)

bias_code_to_desc <- c(
  "41"="Anti-Gay (Male)",   "42"="Anti-Lesbian",
  "43"="Anti-Lesbian, Gay, Bisexual, or Transgender (Mixed)",
  "44"="Anti-Heterosexual", "45"="Anti-Bisexual",
  "71"="Anti-Transgender",  "72"="Anti-Gender Non-Conforming"
)
```

```{r srs-parser}
parse_srs_master_file <- function(filepath) {
  if (!file.exists(filepath)) {
    warning("SRS Master File not found: ", filepath); return(NULL)
  }
  raw_lines <- readLines(filepath, warn = FALSE)
  ir_lines  <- raw_lines[substr(raw_lines, 1, 2) == "IR"]
  lgbtq_bias_codes <- c("41","42","43","44","45","71","72")
  extract_field <- function(lines, start, end) trimws(substr(lines, start, end))
  bias_codes <- extract_field(ir_lines, 222, 223)
  lgbtq_ir   <- ir_lines[bias_codes %in% lgbtq_bias_codes]
  if (length(lgbtq_ir) == 0) { message("No LGBTQI+ records found."); return(NULL) }

  state_abbr  <- extract_field(lgbtq_ir, 5, 6)
  date_raw    <- extract_field(lgbtq_ir, 26, 33)
  offense_raw <- extract_field(lgbtq_ir, 42, 44)
  bias_raw    <- extract_field(lgbtq_ir, 222, 223)
  inc_dates   <- as.Date(date_raw, format = "%Y%m%d")
  yr   <- as.integer(format(inc_dates, "%Y"))
  mo   <- as.integer(format(inc_dates, "%m"))
  offense_names <- coalesce(nibrs_offense_map[offense_raw],
                            paste0("Code ", offense_raw))

  tibble(
    INCIDENT_ID   = extract_field(lgbtq_ir, 14, 25),
    YEAR          = yr, MONTH = mo,
    MONTH_NAME    = as.character(month(inc_dates, label=TRUE, abbr=TRUE)),
    SEASON        = season_from_month(mo),
    STATE_ABBR    = state_abbr,
    STATE_NAME    = srs_state_lookup$name[match(state_abbr, srs_state_lookup$abbr)],
    BIAS_DESC     = bias_code_to_desc[bias_raw],
    BIAS_CATEGORY = classify_bias(bias_code_to_desc[bias_raw]),
    OFFENSE_NAME  = offense_names,
    IS_VIOLENT    = as.integer(str_detect(offense_names,
                      paste(violent_terms, collapse="|"))),
    SEVERITY_TIER = severity_tier(offense_names),
    DATA_SOURCE   = "SRS_MASTER_FILE_2026"
  )
}
```

```{r srs-append, eval=INCLUDE_2026_SRS}
if (file.exists(SRS_FILE)) {
  srs_data <- parse_srs_master_file(SRS_FILE)
  if (!is.null(srs_data) && nrow(srs_data) > 0) {
    df_full_tagged <- df |> select(all_of(full_cols)) |> mutate(DATA_SOURCE = "FBI_UCR_CSV")
    common_cols    <- intersect(names(df_full_tagged), names(srs_data))
    bind_rows(df_full_tagged |> select(all_of(common_cols)),
              srs_data       |> select(all_of(common_cols))) |>
      write_csv(file.path(DATA_DIR, "lgbtq_hate_crimes_full.csv"))
    cat(sprintf("2026 SRS data appended — %d LGBTQI+ incidents added.\n", nrow(srs_data)))
  }
}
```
