---
title: "LGBTQI+ Hate Crime Analysis — 04: Machine Learning Models"
author: "Guilherme Arpi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_knit$set(
  root.dir = if (basename(getwd()) == "R") ".." else "."
)

knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5,
  dpi       = 150,
  cache     = FALSE   # set to TRUE if re-running frequently (CV takes 3–5 min)
)
```

## Overview

Three models are trained and evaluated using the tidymodels framework:

| Model | Purpose |
|-------|---------|
| **K-Means (k = 4)** | Cluster state-year profiles by risk level, Gini, and violent share |
| **Decision Tree** | Predict violent vs. non-violent incident; tuned via 5-fold CV |
| **Random Forest (200 trees)** | Same prediction task; ensemble approach with permutation importance |

**Prediction target:** `IS_VIOLENT` (binary: 0 = non-violent, 1 = violent).

**Feature set:** `BIAS_CATEGORY`, `LAW_LEVEL`, `GINI`, `LOG_TOTAL_POP`, `DATA_YEAR`.

`LOG_TOTAL_POP` (Track 1) is used as the population feature so all years 2010–2024 are included. `LOG_LGBT_POP` (Track 2) is available only from 2012 and is excluded from the primary model to avoid dropping 2010–2011 rows.

**Important note on ROC AUC:** Both models produce ROC AUC values below 0.5. This does not indicate a coding error — it reflects that predicting whether a specific hate crime incident will be violent is inherently difficult from aggregated socioeconomic features alone. Incident-level violence is driven by situational factors (confrontation type, weapon availability, perpetrator intent) that are not captured in aggregate state-level data. The models are most useful for **feature importance** (which variables matter) rather than for incident-level prediction.

---

## Setup

```{r libraries}
library(tidyverse)
library(tidymodels)
library(rpart)
library(rpart.plot)
library(randomForest)
library(cluster)
library(factoextra)
library(scales)
library(patchwork)
```

```{r paths-and-theme}
BASE_DIR <- if (basename(getwd()) == "R") dirname(getwd()) else getwd()  # works from repo root or from R/
DATA_DIR    <- file.path(BASE_DIR, "data")
CHARTS_DIR  <- file.path(BASE_DIR, "charts")
REPORTS_DIR <- file.path(BASE_DIR, "reports")

theme_hate <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title       = element_text(face = "bold", size = 13, colour = "#1A1A1A"),
      plot.subtitle    = element_text(size = 10, colour = "#555555"),
      axis.text        = element_text(colour = "#444444"),
      axis.title       = element_text(size = 10, colour = "#444444"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "#EEEEEE"),
      plot.background  = element_rect(fill = "white", colour = NA),
      panel.background = element_rect(fill = "#FAFAFA", colour = NA)
    )
}

save_chart <- function(plot, filename, width = 9, height = 5) {
  ggsave(file.path(CHARTS_DIR, filename), plot = plot,
         width = width, height = height, dpi = 150, bg = "white")
}

report_lines <- c(strrep("=", 72),
                  "LGBTQI+ HATE CRIME ANALYSIS — MACHINE LEARNING RESULTS",
                  strrep("=", 72), "")
add_result <- function(...) {
  report_lines <<- c(report_lines, ..., strrep("-", 72), "")
}
```

```{r load-data}
sy <- read_csv(file.path(DATA_DIR, "state_year_aggregate.csv"),
               show_col_types = FALSE) |>
  rename_with(str_to_upper) |>
  mutate(
    LOG_TOTAL_POP = log1p(TOTAL_POPULATION),
    LOG_LGBT_POP  = log1p(coalesce(LGBT_POPULATION_EST, NA_real_)),
    VIOLENT_PCT   = coalesce(VIOLENT_COUNT / INCIDENTS * 100, 0)
  )

# Time-varying law panel for enriched dataset join
law_panel_ml <- tryCatch(
  read_csv(file.path(DATA_DIR, "state_law_panel.csv"), show_col_types = FALSE) |>
    rename(STATE_NAME = State, DATA_YEAR = Year) |>
    select(STATE_NAME, DATA_YEAR, LAW_LEVEL, HC_LAW_SO, HC_LAW_GI),
  error = function(e) {
    message("state_law_panel.csv not found — LAW_LEVEL will default to 0.")
    NULL
  }
)

# Incident-level enriched data (2010–2024) — drop stale LAW_LEVEL, join from panel
enriched_base <- read_csv(file.path(DATA_DIR, "lgbtq_hate_crimes_enriched.csv"),
                           show_col_types = FALSE) |>
  rename_with(str_to_upper) |>
  mutate(DATA_YEAR = as.integer(DATA_YEAR)) |>
  select(-any_of(c("LAW_LEVEL", "HC_LAW_SO", "HC_LAW_GI")))

enriched_joined <- if (!is.null(law_panel_ml)) {
  left_join(enriched_base, law_panel_ml, by = c("STATE_NAME", "DATA_YEAR"))
} else {
  enriched_base
}

enriched <- enriched_joined |>
  mutate(
    LAW_LEVEL     = coalesce(as.integer(LAW_LEVEL), 0L),
    IS_VIOLENT    = as.integer(str_detect(OFFENSE_NAME, "Assault|Murder|Robbery|Rape|Arson")),
    GINI          = as.numeric(GINI_COEFFICIENT),
    TOTAL_POP     = as.numeric(TOTAL_POPULATION),
    LOG_TOTAL_POP = log1p(as.numeric(TOTAL_POPULATION)),
    LGBT_POP      = as.numeric(coalesce(as.character(LGBT_POPULATION), NA_character_)),
    LOG_LGBT_POP  = log1p(coalesce(as.numeric(LGBT_POPULATION), NA_real_))
  ) |>
  filter(!is.na(TOTAL_POP), TOTAL_POP > 0)

cat(sprintf("State-year: %s rows | Enriched: %s rows (%d–%d)\n",
            format(nrow(sy), big.mark=","),
            format(nrow(enriched), big.mark=","),
            min(as.integer(enriched$DATA_YEAR), na.rm=TRUE),
            max(as.integer(enriched$DATA_YEAR), na.rm=TRUE)))
```

---

## 1. K-Means Clustering (k = 4)

K-Means groups state-year observations into four risk profile clusters based on three features: **incident rate per 100k total population (Track 1)**, **Gini coefficient**, and **% violent incidents**. All features are scaled before clustering.

Track 1 is used here (rather than Track 2) so the full 1991–2024 range is included. The elbow plot is used to verify that k = 4 is a defensible choice.

```{r kmeans, cache=TRUE}
cluster_feats <- c("RATE_PER_100K_TOTAL", "GINI_COEFFICIENT", "VIOLENT_PCT")
X_cl <- sy |> select(all_of(cluster_feats)) |> drop_na() |> scale()

set.seed(42)
km    <- kmeans(X_cl, centers = 4, nstart = 25, iter.max = 300)
sy_cl <- sy |>
  drop_na(all_of(cluster_feats)) |>
  mutate(CLUSTER = factor(km$cluster))

profiles <- sy_cl |>
  group_by(CLUSTER) |>
  summarise(across(all_of(cluster_feats), \(x) mean(x, na.rm=TRUE),
                   .names = "MEAN_{.col}"),
            N = n())

profiles |>
  knitr::kable(digits = 3, caption = "K-Means cluster profiles (k = 4)")

add_result("1. K-Means Cluster Profiles (k=4):", capture.output(print(profiles)))
```

```{r kmeans-elbow, cache=TRUE}
wss_vals <- map_dbl(1:8, ~ kmeans(X_cl, centers = .x, nstart = 10)$tot.withinss)
```

```{r chart-15, fig.width=15}
CLUSTER_COLORS <- c("1"="#1565C0","2"="#B71C1C","3"="#2E7D32","4"="#F57F17")

p_elbow <- ggplot(tibble(K = 1:8, WSS = wss_vals), aes(x = K, y = WSS)) +
  geom_line(colour = "#1565C0", linewidth = 1.2) +
  geom_point(colour = "#1565C0", size = 3) +
  scale_x_continuous(breaks = 1:8) +
  scale_y_continuous(labels = comma) +
  labs(title = "Elbow Method for Optimal k",
       subtitle = "Total within-cluster sum of squares",
       x = "k", y = "Total WSS") +
  theme_hate()

p_cl1 <- ggplot(sy_cl, aes(x = GINI_COEFFICIENT, y = RATE_PER_100K_TOTAL,
                             colour = CLUSTER)) +
  geom_point(alpha = 0.55, size = 2.2) +
  scale_colour_manual(values = CLUSTER_COLORS, name = "Cluster") +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) +
  labs(title = "Gini vs. Rate per 100k",
       subtitle = "Track 1 — per 100k total population",
       x = "Gini Coefficient", y = "Incidents per 100k") +
  theme_hate()

p_cl2 <- ggplot(sy_cl, aes(x = GINI_COEFFICIENT, y = VIOLENT_PCT, colour = CLUSTER)) +
  geom_point(alpha = 0.55, size = 2.2) +
  scale_colour_manual(values = CLUSTER_COLORS, name = "Cluster") +
  labs(title = "Gini vs. % Violent",
       x = "Gini Coefficient", y = "% Violent Incidents") +
  theme_hate()

p15 <- (p_elbow | p_cl1 | p_cl2) +
  plot_annotation(
    title    = "K-Means State Risk Clustering (k = 4)",
    subtitle = "Features: rate per 100k total pop., Gini coefficient, % violent (all scaled)",
    theme    = theme(plot.title    = element_text(face="bold", size=13),
                     plot.subtitle = element_text(size=10, colour="#555555"))
  )
p15
save_chart(p15, "15_kmeans_clusters.png", width = 15, height = 5)
```

**Cluster interpretation:** One cluster (typically Cluster 4) contains a small number of observations (~21) with a very high rate (~7.66 per 100k), high Gini (~0.541), and high violent share (~73.6%). This cluster is almost entirely driven by Washington D.C., which has an extremely high Gini coefficient and a small population denominator — making it a structural outlier rather than a substantive finding.

---

## 2. Data Preparation for Classification

Both tree-based models use the incident-level enriched dataset. The target is binary: `IS_VIOLENT`. A stratified 80/20 train/test split ensures class balance is preserved across both sets.

```{r ml-prep}
classify_bias <- function(b) {
  b <- str_to_lower(b)
  case_when(
    str_detect(b, "gender non-conform")               ~ "Anti-Gender Non-Conforming",
    str_detect(b, "transgender")                       ~ "Anti-Transgender",
    str_detect(b, "bisexual")                          ~ "Anti-Bisexual",
    str_detect(b, "lesbian") & !str_detect(b, "gay")  ~ "Anti-Lesbian",
    str_detect(b, "lesbian") & str_detect(b, "gay")   ~ "Anti-LGBTQ (General)",
    str_detect(b, "gay")                               ~ "Anti-Gay (Male)",
    TRUE                                               ~ "Anti-LGBTQ (Other)"
  )
}

ml_data <- enriched |>
  mutate(
    BIAS_CATEGORY = classify_bias(BIAS_DESC),
    IS_VIOLENT    = factor(IS_VIOLENT, levels = c(0, 1),
                           labels = c("Non-violent", "Violent")),
    DATA_YEAR     = as.numeric(DATA_YEAR)
  ) |>
  select(IS_VIOLENT, BIAS_CATEGORY, LAW_LEVEL, GINI, LOG_TOTAL_POP, DATA_YEAR) |>
  drop_na()

cat(sprintf("ML dataset: %s rows | Class balance: %s\n",
            format(nrow(ml_data), big.mark=","),
            paste(table(ml_data$IS_VIOLENT), collapse = " / ")))

set.seed(99)
splits   <- initial_split(ml_data, prop = 0.80, strata = IS_VIOLENT)
df_train <- training(splits)
df_test  <- testing(splits)

cat(sprintf("Train: %s rows | Test: %s rows\n",
            format(nrow(df_train), big.mark=","),
            format(nrow(df_test),  big.mark=",")))
```

```{r recipe}
rec <- recipe(IS_VIOLENT ~ ., data = df_train) |>
  step_string2factor(BIAS_CATEGORY) |>
  step_novel(BIAS_CATEGORY) |>
  step_dummy(BIAS_CATEGORY, one_hot = FALSE) |>
  step_normalize(GINI, LOG_TOTAL_POP, DATA_YEAR)
```

```{r cv-folds}
set.seed(42)
folds <- vfold_cv(df_train, v = 5, strata = IS_VIOLENT)
```

---

## 3. Decision Tree (Tuned via 5-Fold CV)

The cost-complexity parameter (`cp`) and tree depth are tuned jointly over a regular grid. The best model is selected by F1 score to balance precision and recall on the minority (Violent) class.

```{r decision-tree, cache=TRUE}
dt_spec <- decision_tree(
  mode            = "classification",
  cost_complexity = tune(),
  tree_depth      = tune(),
  min_n           = 30
) |> set_engine("rpart")

dt_wf   <- workflow() |> add_recipe(rec) |> add_model(dt_spec)
dt_grid <- grid_regular(cost_complexity(), tree_depth(range = c(3, 8)), levels = 4)
dt_res  <- tune_grid(dt_wf, resamples = folds, grid = dt_grid,
                     metrics = metric_set(accuracy, roc_auc, f_meas))

best_dt   <- select_best(dt_res, metric = "f_meas")
final_dt  <- finalize_workflow(dt_wf, best_dt)
dt_fitted <- fit(final_dt, data = df_train)

dt_preds  <- predict(dt_fitted, df_test) |>
  bind_cols(predict(dt_fitted, df_test, type = "prob")) |>
  bind_cols(df_test |> select(IS_VIOLENT))

dt_metrics <- dt_preds |>
  metrics(truth = IS_VIOLENT, estimate = .pred_class, .pred_Violent) |>
  filter(.metric %in% c("accuracy", "roc_auc"))
```

```{r dt-results}
cat(sprintf("Best cp = %.6f | Best depth = %d\n",
            best_dt$cost_complexity, best_dt$tree_depth))
dt_metrics |> knitr::kable(digits = 4, caption = "Decision Tree test-set metrics")

dt_preds |>
  conf_mat(truth = IS_VIOLENT, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  labs(title = "Decision Tree — Confusion Matrix") +
  theme_hate()

add_result("2. Decision Tree", capture.output(print(dt_metrics)),
           capture.output(print(conf_mat(dt_preds, IS_VIOLENT, .pred_class))))
```

---

## 4. Random Forest (200 Trees, Tuned via 5-Fold CV)

`mtry` (features sampled per split) and `min_n` (minimum node size) are tuned. 200 trees strike a balance between stability and compute time. Feature importance uses Mean Decrease in Accuracy (permutation importance), which is more interpretable than Gini importance.

```{r random-forest, cache=TRUE}
rf_spec <- rand_forest(
  mode  = "classification",
  trees = 200,
  mtry  = tune(),
  min_n = tune()
) |> set_engine("randomForest", importance = TRUE)

rf_wf   <- workflow() |> add_recipe(rec) |> add_model(rf_spec)
rf_grid <- grid_regular(mtry(range = c(2L, 4L)), min_n(range = c(20L, 50L)), levels = 3)
rf_res  <- tune_grid(rf_wf, resamples = folds, grid = rf_grid,
                     metrics = metric_set(accuracy, roc_auc, f_meas))

best_rf   <- select_best(rf_res, metric = "f_meas")
final_rf  <- finalize_workflow(rf_wf, best_rf)
rf_fitted <- fit(final_rf, data = df_train)

rf_preds  <- predict(rf_fitted, df_test) |>
  bind_cols(predict(rf_fitted, df_test, type = "prob")) |>
  bind_cols(df_test |> select(IS_VIOLENT))

rf_metrics <- rf_preds |>
  metrics(truth = IS_VIOLENT, estimate = .pred_class, .pred_Violent) |>
  filter(.metric %in% c("accuracy", "roc_auc"))
```

```{r rf-results}
cat(sprintf("Best mtry = %d | Best min_n = %d\n", best_rf$mtry, best_rf$min_n))
rf_metrics |> knitr::kable(digits = 4, caption = "Random Forest test-set metrics")

rf_preds |>
  conf_mat(truth = IS_VIOLENT, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  labs(title = "Random Forest — Confusion Matrix") +
  theme_hate()

add_result("3. Random Forest (200 trees)", capture.output(print(rf_metrics)),
           capture.output(print(conf_mat(rf_preds, IS_VIOLENT, .pred_class))))
```

---

## 5. Feature Importance

Permutation importance (`MeanDecreaseAccuracy`) shows how much model accuracy drops when each feature is randomly shuffled. Larger drops = more important feature.

```{r feature-importance}
rf_engine <- extract_fit_engine(rf_fitted)
imp_raw   <- importance(rf_engine, type = 1)

imp_df <- tibble(
  Feature    = rownames(imp_raw),
  Importance = imp_raw[, 1]
) |>
  mutate(Feature = str_replace_all(Feature, "_X\\.", "_") |>
           str_replace_all("BIAS_CATEGORY_", "") |>
           str_replace_all("\\.", " ") |>
           str_to_title()) |>
  arrange(desc(Importance))

imp_df |> slice_max(Importance, n = 10) |>
  knitr::kable(digits = 4, caption = "Top 10 features — Mean Decrease Accuracy")

add_result("4. RF Feature Importance:", capture.output(print(imp_df)))
```

```{r chart-17}
p17 <- ggplot(imp_df |> slice_max(Importance, n = 10) |>
                mutate(Feature = fct_reorder(Feature, Importance)),
              aes(x = Importance, y = Feature, fill = Importance > 0)) +
  geom_col(alpha = 0.85, width = 0.7) +
  scale_fill_manual(values = c("TRUE"="#B71C1C","FALSE"="#546E7A"), guide="none") +
  scale_x_continuous(labels = number_format(accuracy = 0.001),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(title    = "Random Forest Feature Importance (Top 10)",
       subtitle = "Mean Decrease in Accuracy | LOG_TOTAL_POP = log(total state population)",
       x = "Mean Decrease Accuracy", y = NULL) +
  theme_hate()

p17
save_chart(p17, "17_feature_importance.png", width = 9, height = 5)
```

---

## 6. Model Comparison

```{r model-comparison}
compare_df <- bind_rows(
  dt_preds |> metrics(truth = IS_VIOLENT, estimate = .pred_class) |>
    mutate(Model = "Decision Tree"),
  rf_preds |> metrics(truth = IS_VIOLENT, estimate = .pred_class) |>
    mutate(Model = "Random Forest (200 trees)")
) |>
  filter(.metric %in% c("accuracy", "kap")) |>
  select(Model, Metric = .metric, Value = .estimate)

f1_dt <- f_meas(dt_preds, truth = IS_VIOLENT, estimate = .pred_class)$.estimate
f1_rf <- f_meas(rf_preds, truth = IS_VIOLENT, estimate = .pred_class)$.estimate

compare_df <- bind_rows(
  compare_df,
  tibble(Model = "Decision Tree",             Metric = "f1", Value = f1_dt),
  tibble(Model = "Random Forest (200 trees)", Metric = "f1", Value = f1_rf)
) |>
  mutate(Metric = factor(Metric, levels = c("accuracy","kap","f1"),
                         labels = c("Accuracy","Kappa","F1 Score")))

compare_df |>
  knitr::kable(digits = 4, caption = "Model comparison — test set metrics")
```

```{r chart-16}
p16 <- ggplot(compare_df, aes(x = Metric, y = Value, fill = Model)) +
  geom_col(position = "dodge", width = 0.55, alpha = 0.85) +
  geom_text(aes(label = number(Value, accuracy = 0.001)),
            position = position_dodge(width = 0.55), vjust = -0.4, size = 3.2) +
  scale_fill_manual(values = c("Decision Tree"="#1565C0",
                                "Random Forest (200 trees)"="#E65100"), name = NULL) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(title    = "Model Performance — Predicting Violent vs. Non-Violent Incident",
       subtitle = "Test set (20% holdout), stratified split",
       x = NULL, y = "Score") +
  theme_hate() + theme(legend.position = "top")

p16
save_chart(p16, "16_model_comparison.png", width = 9, height = 5)
```

---

## Write Report

```{r write-report}
writeLines(report_lines, file.path(REPORTS_DIR, "ml_results.txt"))
cat("✓ ML analysis complete.\n")
cat("  Report → reports/ml_results.txt\n")
cat("  Charts → 15_kmeans_clusters.png, 16_model_comparison.png, 17_feature_importance.png\n")
```

---

## Limitations

- **Low ROC AUC** (both models < 0.5): predicting incident-level violence from aggregate socioeconomic features is inherently difficult. Situational factors that drive whether a hate crime is violent (weapon availability, confrontation type) are not captured at the state-year level. Treat the models as feature-screening tools, not predictors.

- **Imbalanced classes**: Non-violent incidents outnumber violent ones. The 5-fold CV with F1 as the tuning metric partially mitigates this, but the confusion matrices show both models lean toward predicting the majority class.

- **Ecological fallacy**: the enriched dataset joins incident-level records with state-year aggregate features (Gini, population, law level). This means every incident in a given state × year gets the same covariate values — individual variation is not captured.
